Optimizing and Modeling Dynamics in Networks

### 2013-07-23 Compiled/chapter2/chapter2.tex
Full compiled

\newcommand{\CLASSINPUTinnersidemargin}{0.7in}
\newcommand{\CLASSINPUToutersidemargin}{0.7in}
\documentclass{article}

@@

\title{Optimizing and Modeling Dynamics in Networks}
\author{{\sc Ibrahim Matta} \\
Computer Science Department \\
Boston University \\
Boston, MA 02215, USA}
\date{\today}							% Activate to display a given date or no date

@@

\tableofcontents

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering 
   \includegraphics[width=4in]{gas-control-loop} 
   \caption{The gas control loop}
   \label{fig:gas-control-loop}
\end{figure}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{net-control-loop.jpg} 
   \caption{The network control loop}
   \label{fig:net-control-loop}
\end{figure}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2in]{concave-utility.jpg} 
   \caption{Concave utility function}
   \label{fig:concave-utility}
\end{figure}

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2in]{concave-function.jpg} 
   \caption{Concave function. 
   A function $f(.)$ is said to be concave if $f(\alpha x_1 + (1-\alpha)x_2) \geq \alpha f(x_1) + (1-\alpha) f(x_2)$,
   i.e., for any two points $x_1$ and $x_2$, the straight line  that connects $f(x_1)$ and $f(x_2)$ is always below
   or equal to the function $f(.)$ itself. 
   Note that a differentiable concave function has a maximum value at some point $x_{max}$, and 
   that the derivative $\dot{f}(x_{max}) = 0.$
   A strictly concave function would have a strict inequality, whereas a convex function has a cup-like shape and
    has a minimum instead.}
   \label{fig:concave-function}
\end{figure}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=5in]{net-model.jpg} 
   \caption{A network model}
   \label{fig:net-model}
\end{figure}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=1.5in]{convex-set.jpg} 
   \caption{A convex set. A convex set intuitively means that any linear combination of any two points located on the boundary of the region, which is formed by the linear inequalities, lies within the region itself. }
   \label{fig:convex-set}
\end{figure}

@@

Maximizing this function results in maximizing the total weighted throughput for all users. As a special case, for unit weights, the network optimization problem maximizes the total throughput through the network. This might seem to fly in the face of what we think is fair! Consider the following simple example (see Figure~\ref{fig:greedy-alloc}):
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{greedy-alloc.jpg} 
   \caption{Greedy network allocation}
   \label{fig:greedy-alloc}
\end{figure}
given both links have capacities of 6 units, the total throughput allocated to all users is the total network capacity of 12 units. This can be achieved by allocating 6 units of capacity to each of the 1-link flows (users): the ``red" user and the ``blue" user, leaving the 2-link (``green") flow with no capacity allocated to its user. That does not seem ``fair"! A different function $f$ would allocate rates to users differently and so it would provide a different notion of fairness.

@@

Intuitively, max-min fairness means that  resources (link capacities) are allocated to users (flows) so that
the allocation is: 
\begin{enumerate}
\item {\em fair:} all users get equal share of a link, as long as users have demand that would fully consume their share, and
\item {\em efficient:} each link is utilized to the maximum load possible.
\end{enumerate}
In other words, if a user cannot consume its equal share of a link, then the excess capacity must be (recursively) allocated equally among high-demanding users. So, the final outcome is that low-demanding users get exactly what they need, while high-demanding users get equal allocations. Consider the following multi-link network example 
(see Figure~\ref{fig:max-min-alloc}):
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{max-min-fair.jpg} 
   \caption{Max-min fair capacity allocation}
   \label{fig:max-min-alloc}
\end{figure}
all links have capacities of 150 units and we assume elastic traffic sources, i.e., sources that would consume all what they can get. We start with the first (left-most) link since it is used by most users so it is the most loaded one. Each flow using that link gets allocated an equal share of $150/3=50$ units. Proceeding to the next loaded link, the middle one, each of its two flows should get an equal share of 75, however flow $F_3$ is limited by its first link to 50 units of throughput. Thus, flow $F_4$ gets the left-over from $F_3$ to a total allocation of $75+25=100$. The right-most link, at capacity of 150, does not limit the throughput  of $F_4$, which ends up using only 100 units of that link, leaving 50 unused. At the end of this process, we say that the max-min fair allocation  vector is $x^{\rm T} =(50,\; 50,\; 50,\; 100)$.

@@

\end{itemize}
In most of our treatment, we consider the simplest kind of models that are deterministic, time-invariant, continuous-time, and linear. In modeling a controlled system, we characterize the relationships among system variables as a function of time, i.e., dynamic equations. See Figure~\ref{fig:system-model} where functions $f$ and $h$ are generally non-linear functions. 
The function $f$ models the evolution of the state of the system $x$ as 
a function of the current state and system's input $u$.
The function $h$ yields system's output $y$ as a function of the current state and input values.
As we will see later, for mathematical tractability, we often linearize dynamic non-linear models or 
we only consider operation in a linear regime, for example, we ignore non-linearity when 
the buffer goes empty or hits full capacity.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=3in]{system-model.jpg} % requires the graphicx package
   \caption{Typical system model}
   \label{fig:system-model}
\end{figure}

@@

Many analytical studies considered the network system composed of TCP sources over a network of queues that employ a certain queue management policy. Examples of TCP variants include Reno, SACK \cite{Fall:1996}, NewReno, Vegas \cite{vegas:1995}, FAST \cite{fast:2006}, etc. Examples of queue management policies include Drop Tail, RED \cite{red:1993}, REM \cite{REM:2001}, PI \cite{PI:2000}, etc. One of the most widely studied instantiation is that of TCP sources over a RED bottleneck queue --- see Figure~\ref{fig:TCP-RED-diagram}.
We start by modeling the dynamic behavior of a TCP source, i.e., 
the time-dependent relationship between its transmission window (or sending rate) and 
its observed loss rate or delay (price).
We do so for both TCP Reno and Vegas versions, and also deduce their utilities.
Then, we model the buffering process inside the network (more precisely, the bottleneck queue),
assuming a linear regime (i.e., ignoring non-linearities due to the buffer becoming empty or full).
Using the average buffer length, we model the dynamic behavior of RED and how 
it generates packet losses (or markings) as indication of congestion (price).
This overall model represents the closed-loop feedback system shown in 
the block diagram of Figure~\ref{fig:TCP-RED-diagram}, 
which can then be analyzed using control-theoretic techniques.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=4in]{tcp-red.jpg} % requires the graphicx package
   \caption{TCP Reno over RED feedback control system}
   \label{fig:TCP-RED-diagram}
\end{figure}

@@

\paragraph{Modeling RED:}
Let us now consider the modeling of the buffer and associated RED queue management algorithm \cite{red:1993}.
Figure~\ref{fig:red} shows how RED tries to avoid congestion by dropping (or marking) packets with probability $p_c$ 
as a (non-linear) function of the average queue length $v$.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=2in]{red.jpg} % requires the graphicx package
   \caption{RED dropping (or marking) function}
   \label{fig:red}
\end{figure}
First, we model the evolution of the queue length $b(t)$ as 
a function of the total input rate, $y(t) = \sum x_s(t)$, and (bottleneck) link capacity, $C$:
\begin{eqnarray*}
\dot{b}(t) =  y(t) - C
\end{eqnarray*}
Denoting by $v(t)$, the Exponential Weighted Moving Average (EWMA) of the queue length:
\begin{eqnarray*}
v(t+\delta) = (1 - \alpha) v(t) + \alpha b(t)
\end{eqnarray*}
\begin{eqnarray*}
v(t+\delta) - v(t) = \alpha (b(t) - v(t))
\end{eqnarray*}
Given $v(t)$ gets updated at the link rate, i.e. $\delta = \frac{1}{C}$,
and $\dot{v}(t) \approx \frac{v(t+\delta) - v(t)}{\delta}$, we have:
\begin{eqnarray*}
\dot{v}(t) = \alpha C \; (b(t) - v(t))
\end{eqnarray*}
This last equation represents the dynamic model of RED averaging, which in turn determines the price $p_c(t)$ that users experience.

@@

To this end, a technique called {\em linearization} can prove more tractable
where the non-linear system is approximated by a set of {\em linear} equations around a single operating point (state).
See Figure \ref{fig:linearize}.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=3in]{linearize.jpg} % requires the graphicx package
   \caption{Linearization}
   \label{fig:linearize}
\end{figure}
With linearization, we become concerned with {\em local stability} and study perturbations around the operating point
using standard (linear) control theory. 
By local stability, we mean that if the system is perturbed within a small region around the operating point 
then the system will converge and stabilize back to that point.
This is in contrast to {\em global stability} where the original (non-linear) system is shown to 
converge from {\em any} starting state.
To linearize the non-linear system around an operating point,
the basic idea is to expand the non-linear differential equation into 
a Taylor series around that point and then ignore high-order terms.

@@

Consider the system in Figure~\ref{fig:vegas-control} where a controller $G_c$ is used to match
the queue length $b(t)$ to a target $B_r$ by controlling the input window or sending rate $x(t)$.
The output rate from the queue is denoted by $c(t)$.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=2.5in]{vegas-control.jpg} % requires the graphicx package
   \caption{Vegas-like system}
   \label{fig:vegas-control}
\end{figure}
The goal is to first write down the differential equations that model the different components of the system,
then instead of solving the equations in the time domain, we will transform them to the Laplace domain and
analyze the stability of the system algebraically.

@@

Figure~\ref{fig:vegas-BD} shows the system using its transfer functions and their input/output flows,
where $G_0 = \frac{1}{s}$.
This is called the {\em block diagram} and provides a powerful pictorial tool.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=4in]{vegas-BD.jpg} % requires the graphicx package
   \caption{Block diagram of Vegas-like system}
   \label{fig:vegas-BD}
\end{figure} 
From the block diagram, one can write the algebraic equation of the output in terms of the input(s).
Dropping the ``s" parameter for convenience:
\begin{eqnarray*}
\frac{(B_r - B) G_c - C}{s} = B
\end{eqnarray*}
Rearranging, we get:
\begin{eqnarray}
B = \frac{G_c}{s + G_c} B_r - \frac{1}{s + G_c} C
\label{eqn:vegas-TF}
\end{eqnarray}
Note that the system has two inputs: $B_r(s)$ and $C(s)$, subjected to two transfer functions,
$\frac{G_c(s)}{s + G_c(s)}$ and $\frac{-1}{s + G_c(s)}$, respectively, and 
adding their responses we obtain the output $B(s)$.

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=1.6in]{overdamped.jpg} 
   \caption{Overdamped response}
   \label{fig:overdamped}
\end{figure}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2.5in]{undamped.jpg} 
   \caption{Undamped response}
   \label{fig:undamped}
\end{figure}

@@

Note that if the poles are complex conjugates and strictly in the left-hand side 
of the s-plane, the system is stable as oscillations in response to impulse input decay over time,
and we say that the system exhibits {\em underdamped response}.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=4in]{pole-loc.jpg} % requires the graphicx package
   \caption{Time response depending on pole location}
   \label{fig:pole-loc}
\end{figure}

@@

\begin{figure}[htbp]
   \centering
   \includegraphics[width=4in]{perf-spec.jpg} % requires the graphicx package
   \caption{Performance Metrics}
   \label{fig:transient}
\end{figure}

@@

\begin{figure}[htbp]
   \centering
   \includegraphics[width=4in]{cauchy.jpg} % requires the graphicx package
   \caption{Cauchy's principle}
   \label{fig:cauchy}
\end{figure}

The Nyquist method applies the Cauchy's principle as follows.
Let's say we want to analyze the stability of a closed-loop control system
whose forward transfer function is $G(s)$ and its feedback transfer function is $H(s)$---see Figure \ref{fig:closed-loop}.
Then, the closed-loop transfer function is given by $\frac{G(s)}{1+G(s) H(s)}$,
where $G(s)H(s)$ is referred to as the {\em open-loop} transfer function.
The characteristic equation is given by: $F(s) = 1 + G(s) H(s) = 0$.
Observe that the zeros of $F(s)$ are the closed-loop poles,
and the poles of $F(s)$ are the poles of $G(s)H(s)$ (so-called ``open-loop" poles).
\begin{figure}[htbp]
   \centering
   \includegraphics[width=3in]{closed-loop.jpg} % requires the graphicx package
   \caption{Typical closed-loop control system}
   \label{fig:closed-loop}
\end{figure}

By taking the s-contour to be around the right-side (i.e. unstable side) of the s-plane
(see Figure \ref{fig:s-contour}),
and noting the number of {\em unstable} open-loop poles $P$
and the number of encirclements $N$ around the origin in the $F(s)$-plane, we determine 
the number of {\em unstable} zeros $Z$ of $F(s)$, i.e. number of unstable closed-loop poles,
using Cauchy's relationship: $Z = P + N$.
If $P = 0$ and $N = 0$, then $Z = 0$ implies that there are no {\em unstable} closed-loop poles
and so the closed-loop system is stable.\footnote{
Note that a pole on the imaginary axis is not considered unstable and
the s-contour avoids such a pole and we show it as a small circle around it.
}
\begin{figure}[htbp]
   \centering
   \includegraphics[width=3in]{s-contour.jpg} % requires the graphicx package
   \caption{Contour around the unstable right-side of the s-plane}
   \label{fig:s-contour}
\end{figure}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=3in]{ex-nyquist.jpg} 
   \caption{Example showing the effect of feedback delay}
   \label{fig:ex-nyquist}
\end{figure}

@@

So far, we assumed routes taken by flows to be static. In general, routes may also be adapted based on feedback on link prices (reflecting load, delay, etc.), 
albeit over a longer timescale of minutes, hours or even days compared to that of milliseconds for sending rate adaptation.
Figure \ref{fig:block-routing} shows a block diagram that includes both route and sending rate adaptation.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=5in]{block-routing.jpg} % requires the graphicx package
   \caption{Block diagram with both flow and routing control}
   \label{fig:block-routing}
\end{figure}

@@

\begin{figure}[htbp]
   \centering
   \includegraphics[width=4in]{routing-conv.jpg} % requires the graphicx package
   \caption{Convergence}
   \label{fig:routing-conv}
\end{figure}

@@

\begin{figure}[htbp]
   \centering
   \includegraphics[width=3in]{contractive.jpg} % requires the graphicx package
   \caption{Contractive mapping}
   \label{fig:contractive}
\end{figure}

@@

\begin{figure}[htbp]
   \centering
   \includegraphics[width=3in]{non-contractive.jpg} % requires the graphicx package
   \caption{Divergent mapping}
   \label{fig:non-contractive}
\end{figure}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{case1-block.jpg} 
   \caption{Block diagram of the link sharing model}
   \label{fig:case1-block}
\end{figure}

@@

  \begin{figure}[htbp] %  figure placement: here, top, bottom, or page
     \centering
     \includegraphics[width=4in]{case2-tunnel.jpg} 
     \caption{Soft bandwidth-guaranteed tunnel}
     \label{fig:case2-tunnel}
  \end{figure}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{tunnel-block1} 
\caption{Block diagram of the elastic-tunnel model} 
\label{fig:block1}
\end{figure}

@@

\begin{figure}[tbp]
   \includegraphics[width=5in]{tunnel-block} 
\caption{Elastic-tunnel model (with feedback delay)} 
\label{fig:block2}
\end{figure}

@@

\begin{figure*}[htp]
\centerline{
  % Requires \usepackage{graphicx}
  \includegraphics[width=0.7\textwidth]{blockDiagram}
  }
   \caption{System block diagram}
   \label{fig:sol-block}
\end{figure*}

>>>>

%\newcommand{\CLASSINPUTinnersidemargin}{0.7in}
%\newcommand{\CLASSINPUToutersidemargin}{0.7in}
\documentclass{article}

@@

\title{Optimizing and Modeling Dynamics in Networks}
\author{{\sc Ibrahim Matta} \\
Computer Science Department \\
Boston University \\
Boston, MA 02215, USA}
\date{}							% Activate to display a given date or no date

@@

%\tableofcontents

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering 
   \includegraphics[width=4in]{figures/gas-control-loop} 
   \caption{The gas control loop}
   \label{fig:gas-control-loop}
\end{figure}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{figures/net-control-loop.jpg} 
   \caption{The network control loop}
   \label{fig:net-control-loop}
\end{figure}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2in]{figures/concave-utility.jpg} 
   \caption{Concave utility function}
   \label{fig:concave-utility}
\end{figure}

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2in]{figures/concave-function.jpg} 
   \caption{Concave function. 
   A function $f(.)$ is said to be concave if $f(\alpha x_1 + (1-\alpha)x_2) \geq \alpha f(x_1) + (1-\alpha) f(x_2)$,
   i.e., for any two points $x_1$ and $x_2$, the straight line  that connects $f(x_1)$ and $f(x_2)$ is always below
   or equal to the function $f(.)$ itself. 
   Note that a differentiable concave function has a maximum value at some point $x_{max}$, and 
   that the derivative $\dot{f}(x_{max}) = 0.$
   A strictly concave function would have a strict inequality, whereas a convex function has a cup-like shape and
    has a minimum instead.}
   \label{fig:concave-function}
\end{figure}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=5in]{figures/net-model.jpg} 
   \caption{A network model}
   \label{fig:net-model}
\end{figure}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=1.5in]{figures/convex-set.jpg} 
   \caption{A convex set. A convex set intuitively means that any linear combination of any two points located on the boundary of the region, which is formed by the linear inequalities, lies within the region itself. }
   \label{fig:convex-set}
\end{figure}

@@

Maximizing this function results in maximizing the total weighted throughput for all users. As a special case, for unit weights, the network optimization problem maximizes the total throughput through the network. This might seem to fly in the face of what we think is fair! Consider the following simple example (see Figure~\ref{fig:greedy-alloc}):
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{figures/greedy-alloc.jpg} 
   \caption{Greedy network allocation}
   \label{fig:greedy-alloc}
\end{figure}
given both links have capacities of 6 units, the total throughput allocated to all users is the total network capacity of 12 units. This can be achieved by allocating 6 units of capacity to each of the 1-link flows (users): the ``red" user and the ``blue" user, leaving the 2-link (``green") flow with no capacity allocated to its user. That does not seem ``fair"! A different function $f$ would allocate rates to users differently and so it would provide a different notion of fairness.

@@

Intuitively, max-min fairness means that  resources (link capacities) are allocated to users (flows) so that
the allocation is: 
\begin{enumerate}
\item {\em fair:} all users get equal share of a link, as long as users have demand that would fully consume their share, and
\item {\em efficient:} each link is utilized to the maximum load possible.
\end{enumerate}
In other words, if a user cannot consume its equal share of a link, then the excess capacity must be (recursively) allocated equally among high-demanding users. So, the final outcome is that low-demanding users get exactly what they need, while high-demanding users get equal allocations. Consider the following multi-link network example 
(see Figure~\ref{fig:max-min-alloc}):
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{figures/max-min-fair.jpg} 
   \caption{Max-min fair capacity allocation}
   \label{fig:max-min-alloc}
\end{figure}
all links have capacities of 150 units and we assume elastic traffic sources, i.e., sources that would consume all what they can get. We start with the first (left-most) link since it is used by most users so it is the most loaded one. Each flow using that link gets allocated an equal share of $150/3=50$ units. Proceeding to the next loaded link, the middle one, each of its two flows should get an equal share of 75, however flow $F_3$ is limited by its first link to 50 units of throughput. Thus, flow $F_4$ gets the left-over from $F_3$ to a total allocation of $75+25=100$. The right-most link, at capacity of 150, does not limit the throughput  of $F_4$, which ends up using only 100 units of that link, leaving 50 unused. At the end of this process, we say that the max-min fair allocation  vector is $x^{\rm T} =(50,\; 50,\; 50,\; 100)$.

@@

\end{itemize}
In most of our treatment, we consider the simplest kind of models that are deterministic, time-invariant, continuous-time, and linear. In modeling a controlled system, we characterize the relationships among system variables as a function of time, i.e., dynamic equations. See Figure~\ref{fig:system-model} where functions $f$ and $h$ are generally non-linear functions. 
The function $f$ models the evolution of the state of the system $x$ as 
a function of the current state and system's input $u$.
The function $h$ yields system's output $y$ as a function of the current state and input values.
As we will see later, for mathematical tractability, we often linearize dynamic non-linear models or 
we only consider operation in a linear regime, for example, we ignore non-linearity when 
the buffer goes empty or hits full capacity.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=3in]{figures/system-model.jpg} % requires the graphicx package
   \caption{Typical system model}
   \label{fig:system-model}
\end{figure}

@@

Many analytical studies considered the network system composed of TCP sources over a network of queues that employ a certain queue management policy. Examples of TCP variants include Reno, SACK \cite{Fall:1996}, NewReno, Vegas \cite{vegas:1995}, FAST \cite{fast:2006}, etc. Examples of queue management policies include Drop Tail, RED \cite{red:1993}, REM \cite{REM:2001}, PI \cite{PI:2000}, etc. One of the most widely studied instantiation is that of TCP sources over a RED bottleneck queue --- see Figure~\ref{fig:TCP-RED-diagram}.
We start by modeling the dynamic behavior of a TCP source, i.e., 
the time-dependent relationship between its transmission window (or sending rate) and 
its observed loss rate or delay (price).
We do so for both TCP Reno and Vegas versions, and also deduce their utilities.
Then, we model the buffering process inside the network (more precisely, the bottleneck queue),
assuming a linear regime (i.e., ignoring non-linearities due to the buffer becoming empty or full).
Using the average buffer length, we model the dynamic behavior of RED and how 
it generates packet losses (or markings) as indication of congestion (price).
This overall model represents the closed-loop feedback system shown in 
the block diagram of Figure~\ref{fig:TCP-RED-diagram}, 
which can then be analyzed using control-theoretic techniques.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=4in]{figures/tcp-red.jpg} % requires the graphicx package
   \caption{TCP Reno over RED feedback control system}
   \label{fig:TCP-RED-diagram}
\end{figure}

@@

\paragraph{Modeling RED:}
Let us now consider the modeling of the buffer and associated RED queue management algorithm \cite{red:1993}.
Figure~\ref{fig:red} shows how RED tries to avoid congestion by dropping (or marking) packets with probability $p_c$ 
as a (non-linear) function of the average queue length $v$.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=2in]{figures/red.jpg} % requires the graphicx package
   \caption{RED dropping (or marking) function}
   \label{fig:red}
\end{figure}
First, we model the evolution of the queue length $b(t)$ as 
a function of the total input rate, $y(t) = \sum x_s(t)$, and (bottleneck) link capacity, $C$:
\begin{eqnarray*}
\dot{b}(t) =  y(t) - C
\end{eqnarray*}
Denoting by $v(t)$, the Exponential Weighted Moving Average (EWMA) of the queue length:
\begin{eqnarray*}
v(t+\delta) = (1 - \alpha) v(t) + \alpha b(t)
\end{eqnarray*}
\begin{eqnarray*}
v(t+\delta) - v(t) = \alpha (b(t) - v(t))
\end{eqnarray*}
Given $v(t)$ gets updated at the link rate, i.e. $\delta = \frac{1}{C}$,
and $\dot{v}(t) \approx \frac{v(t+\delta) - v(t)}{\delta}$, we have:
\begin{eqnarray*}
\dot{v}(t) = \alpha C \; (b(t) - v(t))
\end{eqnarray*}
This last equation represents the dynamic model of RED averaging, which in turn determines the price $p_c(t)$ that users experience.

@@

To this end, a technique called {\em linearization} can prove more tractable
where the non-linear system is approximated by a set of {\em linear} equations around a single operating point (state).
See Figure \ref{fig:linearize}.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=3in]{figures/linearize.jpg} % requires the graphicx package
   \caption{Linearization}
   \label{fig:linearize}
\end{figure}
With linearization, we become concerned with {\em local stability} and study perturbations around the operating point
using standard (linear) control theory. 
By local stability, we mean that if the system is perturbed within a small region around the operating point 
then the system will converge and stabilize back to that point.
This is in contrast to {\em global stability} where the original (non-linear) system is shown to 
converge from {\em any} starting state.
To linearize the non-linear system around an operating point,
the basic idea is to expand the non-linear differential equation into 
a Taylor series around that point and then ignore high-order terms.

@@

Consider the system in Figure~\ref{fig:vegas-control} where a controller $G_c$ is used to match
the queue length $b(t)$ to a target $B_r$ by controlling the input window or sending rate $x(t)$.
The output rate from the queue is denoted by $c(t)$.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=2.5in]{figures/vegas-control.jpg} % requires the graphicx package
   \caption{Vegas-like system}
   \label{fig:vegas-control}
\end{figure}
The goal is to first write down the differential equations that model the different components of the system,
then instead of solving the equations in the time domain, we will transform them to the Laplace domain and
analyze the stability of the system algebraically.

@@

Figure~\ref{fig:vegas-BD} shows the system using its transfer functions and their input/output flows,
where $G_0 = \frac{1}{s}$.
This is called the {\em block diagram} and provides a powerful pictorial tool.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=4in]{figures/vegas-BD.jpg} % requires the graphicx package
   \caption{Block diagram of Vegas-like system}
   \label{fig:vegas-BD}
\end{figure} 
From the block diagram, one can write the algebraic equation of the output in terms of the input(s).
Dropping the ``s" parameter for convenience:
\begin{eqnarray*}
\frac{(B_r - B) G_c - C}{s} = B
\end{eqnarray*}
Rearranging, we get:
\begin{eqnarray}
B = \frac{G_c}{s + G_c} B_r - \frac{1}{s + G_c} C
\label{eqn:vegas-TF}
\end{eqnarray}
Note that the system has two inputs: $B_r(s)$ and $C(s)$, subjected to two transfer functions,
$\frac{G_c(s)}{s + G_c(s)}$ and $\frac{-1}{s + G_c(s)}$, respectively, and 
adding their responses we obtain the output $B(s)$.

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=1.6in]{figures/overdamped.jpg} 
   \caption{Overdamped response}
   \label{fig:overdamped}
\end{figure}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2.5in]{figures/undamped.jpg} 
   \caption{Undamped response}
   \label{fig:undamped}
\end{figure}

@@

Note that if the poles are complex conjugates and strictly in the left-hand side 
of the s-plane, the system is stable as oscillations in response to impulse input decay over time,
and we say that the system exhibits {\em underdamped response}.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=4in]{figures/pole-loc.jpg} % requires the graphicx package
   \caption{Time response depending on pole location}
   \label{fig:pole-loc}
\end{figure}

@@

\begin{figure}[htbp]
   \centering
   \includegraphics[width=4in]{figures/perf-spec.jpg} % requires the graphicx package
   \caption{Performance Metrics}
   \label{fig:transient}
\end{figure}

@@

\begin{figure}[htbp]
   \centering
   \includegraphics[width=4in]{figures/cauchy.jpg} % requires the graphicx package
   \caption{Cauchy's principle}
   \label{fig:cauchy}
\end{figure}

The Nyquist method applies the Cauchy's principle as follows.
Let's say we want to analyze the stability of a closed-loop control system
whose forward transfer function is $G(s)$ and its feedback transfer function is $H(s)$---see Figure \ref{fig:closed-loop}.
Then, the closed-loop transfer function is given by $\frac{G(s)}{1+G(s) H(s)}$,
where $G(s)H(s)$ is referred to as the {\em open-loop} transfer function.
The characteristic equation is given by: $F(s) = 1 + G(s) H(s) = 0$.
Observe that the zeros of $F(s)$ are the closed-loop poles,
and the poles of $F(s)$ are the poles of $G(s)H(s)$ (so-called ``open-loop" poles).
\begin{figure}[htbp]
   \centering
   \includegraphics[width=3in]{figures/closed-loop.jpg} % requires the graphicx package
   \caption{Typical closed-loop control system}
   \label{fig:closed-loop}
\end{figure}

By taking the s-contour to be around the right-side (i.e. unstable side) of the s-plane
(see Figure \ref{fig:s-contour}),
and noting the number of {\em unstable} open-loop poles $P$
and the number of encirclements $N$ around the origin in the $F(s)$-plane, we determine 
the number of {\em unstable} zeros $Z$ of $F(s)$, i.e. number of unstable closed-loop poles,
using Cauchy's relationship: $Z = P + N$.
If $P = 0$ and $N = 0$, then $Z = 0$ implies that there are no {\em unstable} closed-loop poles
and so the closed-loop system is stable.\footnote{
Note that a pole on the imaginary axis is not considered unstable and
the s-contour avoids such a pole and we show it as a small circle around it.
}
\begin{figure}[htbp]
   \centering
   \includegraphics[width=3in]{figures/s-contour.jpg} % requires the graphicx package
   \caption{Contour around the unstable right-side of the s-plane}
   \label{fig:s-contour}
\end{figure}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=3in]{figures/ex-nyquist.jpg} 
   \caption{Example showing the effect of feedback delay}
   \label{fig:ex-nyquist}
\end{figure}

@@

So far, we assumed routes taken by flows to be static. In general, routes may also be adapted based on feedback on link prices (reflecting load, delay, etc.), 
albeit over a longer timescale of minutes, hours or even days compared to that of milliseconds for sending rate adaptation.
Figure \ref{fig:block-routing} shows a block diagram that includes both route and sending rate adaptation.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=5in]{figures/block-routing.jpg} % requires the graphicx package
   \caption{Block diagram with both flow and routing control}
   \label{fig:block-routing}
\end{figure}

@@

\begin{figure}[htbp]
   \centering
   \includegraphics[width=4in]{figures/routing-conv.jpg} % requires the graphicx package
   \caption{Convergence}
   \label{fig:routing-conv}
\end{figure}

@@

\begin{figure}[htbp]
   \centering
   \includegraphics[width=3in]{figures/contractive.jpg} % requires the graphicx package
   \caption{Contractive mapping}
   \label{fig:contractive}
\end{figure}

@@

\begin{figure}[htbp]
   \centering
   \includegraphics[width=3in]{figures/non-contractive.jpg} % requires the graphicx package
   \caption{Divergent mapping}
   \label{fig:non-contractive}
\end{figure}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{figures/case1-block.jpg} 
   \caption{Block diagram of the link sharing model}
   \label{fig:case1-block}
\end{figure}

@@

  \begin{figure}[htbp] %  figure placement: here, top, bottom, or page
     \centering
     \includegraphics[width=4in]{figures/case2-tunnel.jpg} 
     \caption{Soft bandwidth-guaranteed tunnel}
     \label{fig:case2-tunnel}
  \end{figure}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{figures/tunnel-block1} 
\caption{Block diagram of the elastic-tunnel model} 
\label{fig:block1}
\end{figure}

@@

\begin{figure}[tbp]
   \includegraphics[width=5in]{figures/tunnel-block} 
\caption{Elastic-tunnel model (with feedback delay)} 
\label{fig:block2}
\end{figure}

@@

\begin{figure*}[htp]
\centerline{
  % Requires \usepackage{graphicx}
  \includegraphics[width=0.7\textwidth]{figures/blockDiagram}
  }
   \caption{System block diagram}
   \label{fig:sol-block}
\end{figure*}
